Combined annotation exploration for SCPCP000006

### 2025-03-17 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

Here we plot for each chromosome arm (x-axis) the number of samples having a `dupli` (y-axis \> 0, in red) or a `loss` (y-axis \< 0, in blue).

@@

Comparing with the profile in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1), we identified important CNV known to play a role in Wilms tumor, including:

@@

- We couldn't detect any *chr4q* loss as described in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).

- We identified 2 losses in *chr5* and *chr6* that do not show up in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
These may be `infercnv` induced false-positive CNV.

Our major concern is the pattern of the *chr1*. 
We do see a tendency of *chr1p* loss and *chr1q* gain but it is not as pronounced as in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). 
*chr1q* gain is however one of the most promising predictive marker of Wilms tumor outcome ([Nelson et al, Current Opinion in Pediatrics (2020)](https://pmc.ncbi.nlm.nih.gov/articles/PMC7919941/)). 
It is however important to notice that the profile shown in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1) derived from SIOP samples that have been pre-treated with chemotherapy, while samples in the dataset are enriched in `upfront resection` samples that haven't been pre-treated, according to the COG protocol. 
So far, we do not know how the pre-operative chemotherapy impact the clonal selection of specific CNV.

@@

As described in [Nelson et al, Current Opinions in Pediatrics (2020)](https://pmc.ncbi.nlm.nih.gov/articles/PMC7919941/), [Phelps et al, Children (2018)](https://pmc.ncbi.nlm.nih.gov/articles/PMC6262554/) and [Zheng et al, Frontiers in Oncology (2023)](https://www.frontiersin.org/journals/oncology/articles/10.3389/fonc.2023.1137346/full), LOH at *chr1p*, *chr11p* and *chr16q* as well as *chr1q* gain are prognostic factors.

@@

As we don't know the effect of the treatment on the selection of cells with CNV, it is difficult to compare our CNV profile with the one in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1). 
We those check the repartition of CNV in samples that have been pre-treated (i.e `Resection post chemotherapy`) with samples that have been removed surgically before any treatment (i.e `Upfront resection`).

@@

Our analysis suggest that there is indeed an enrichment in cells with CNV after chemotherapy, including a tendency of more *chr1q* gain samples after chemotherapy. This could explain some of the differences observed between our mean CNV profile and the one published in [Cresswell et al, bioRxiv (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).

@@

Immune cells are well annotated based on `PTPRC` expression.

@@

Endothelial cells are well annotated based on `VWF` expression.

>>>>

Here we plot for each chromosome arm (x-axis) the number of samples having a `dupli` (y-axis > 0, in red) or a `loss` (y-axis < 0, in blue).

@@

Comparing with the profile in [Cresswell et al. (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1), we identified important CNV known to play a role in Wilms tumor, including:

@@

- We couldn't detect any *chr4q* loss as described in [Cresswell et al. (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).

- We identified 2 losses in *chr5* and *chr6* that do not show up in [Cresswell et al. (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
These may be `infercnv` induced false-positive CNV.

Our major concern is the pattern of the *chr1*. 
We do see a tendency of *chr1p* loss and *chr1q* gain but it is not as pronounced as in [Cresswell et al. (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
*chr1q* gain is however one of the most promising predictive marker of Wilms tumor outcome ([Nelson et al. (2020)](https://pmc.ncbi.nlm.nih.gov/articles/PMC7919941/)).
It is however important to notice that the profile shown in [Cresswell et al. (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1) derived from SIOP samples that have been pre-treated with chemotherapy, while samples in the dataset are enriched in `upfront resection` samples that haven't been pre-treated, according to the COG protocol.
So far, we do not know how the pre-operative chemotherapy impact the clonal selection of specific CNV.

@@

As described in [Nelson et al. (2020)](https://pmc.ncbi.nlm.nih.gov/articles/PMC7919941/), [Phelps et al. (2018)](https://pmc.ncbi.nlm.nih.gov/articles/PMC6262554/) and [Zheng et al. (2023)](https://www.frontiersin.org/journals/oncology/articles/10.3389/fonc.2023.1137346/full), LOH at *chr1p*, *chr11p* and *chr16q* as well as *chr1q* gain are prognostic factors.

@@

As we don't know the effect of the treatment on the selection of cells with CNV, it is difficult to compare our CNV profile with the one in [Cresswell et al. (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).
We those check the repartition of CNV in samples that have been pre-treated (i.e `Resection post chemotherapy`) with samples that have been removed surgically before any treatment (i.e `Upfront resection`).

@@

Our analysis suggest that there is indeed an enrichment in cells with CNV after chemotherapy, including a tendency of more *chr1q* gain samples after chemotherapy. This could explain some of the differences observed between our mean CNV profile and the one published in [Cresswell et al. (2024)](https://www.biorxiv.org/content/10.1101/2024.09.04.610994v1).

@@

Immune cells are well annotated based on *PTPRC* expression.

@@

Endothelial cells are well annotated based on *VWF* expression.
